---
title: "Practical Machine Learning Project"
author: "Agnieszka Chojecka"
date: "`r Sys.Date()`"
output: html_document
---

# Introduction

The purpose of this project is to predict the manner in which six participants performed a barbell exercise using data collected from wearable accelerometers. The outcome variable is **classe**, which has five possible values (A–E).

# Loading Libraries and Data

First, the required libraries are loaded and the training and testing datasets are imported.

```{r, echo=FALSE, message=FALSE, warning=FALSE}
library(caret)
library(randomForest)

trainUrl <- "https://d396qusza40orc.cloudfront.net/predmachlearn/pml-training.csv"
testUrl <- "https://d396qusza40orc.cloudfront.net/predmachlearn/pml-testing.csv"

training <- read.csv(trainUrl)
testing <- read.csv(testUrl)

training$classe <- as.factor(training$classe)
```

The dimensions of both datasets are checked.

```{r, echo=FALSE}
dim(training)
dim(testing)
```

# Data Cleaning

The data were cleaned by:

- removing predictors with near-zero variance,
- removing variables containing missing values,
- removing the first five identification variables.

```{r, echo=FALSE}
nearZero <- nearZeroVar(training)

training <- training[, -nearZero]
testing <- testing[, -nearZero]

# Usuń kolumny z NA
training <- training[, colSums(is.na(training)) == 0]

# Wybierz w testingu te same predyktory co w training (bez klasy)
testing <- testing[, colnames(training)[colnames(training) != "classe"]]

training <- training[, -c(1:5)]
testing <- testing[, -c(1:5)]
```

The dimensions after cleaning are:

```{r, echo=FALSE}
dim(training)
dim(testing)
```

# Data Partitioning

The dataset was randomly split into a training set (70%) and a validation set (30%). The validation set was used to estimate the model's performance on unseen data.

```{r, echo=FALSE}
set.seed(123)

inTrain <- createDataPartition(training$classe,
                               p = 0.7,
                               list = FALSE)

train <- training[inTrain, ]
validation <- training[-inTrain, ]
```

# Model Building

A Random Forest classifier was selected because it performs well for classification problems involving many predictors.

```{r, echo=FALSE}
model <- randomForest(
  classe ~ .,
  data = train,
  ntree = 100
)

model
```

The figure below presents the importance of the predictors used by the Random Forest model.

```{r, fig.width=10, fig.height=8, echo=FALSE}
varImpPlot(model)
```

# Model Evaluation

Predictions were made on the validation dataset and evaluated using a confusion matrix.

```{r, echo=FALSE}
prediction <- predict(model, validation)

cm <- confusionMatrix(prediction,
                      validation$classe)

cm
```

The estimated accuracy of the model is:

```{r, echo=FALSE}
cm$overall["Accuracy"]
```

The estimated out-of-sample error is:

```{r, echo=FALSE}
1 - cm$overall["Accuracy"]
```

The Random Forest model achieved a very high accuracy, indicating that it is appropriate for predicting the exercise classes.

# Prediction on Test Data

Finally, predictions were generated for the 20 observations in the testing dataset.

```{r, echo=FALSE}
finalPrediction <- predict(model, testing)

finalPrediction
```

# Conclusion

After cleaning the data and training a Random Forest model, the classifier achieved a high prediction accuracy on the validation dataset. The trained model was then used to predict the exercise classes for the 20 observations in the testing dataset.